ODGI: understanding pangenome graphs

### 2022-03-18 manuscript/main.tex

%\FIXME{Update the references to the sections.}
\begin{itemize}
\item~\textit{odgi build} constructs the ODGI data model from GFA file (\S\ref{sec:build}).
\item~\textit{odgi view} converts the ODGI data model into GFA file (\S\ref{sec:build}).
\item~\textit{odgi viz} provides a linear visualization of the graph (\S\ref{sec:viz}).
\item~\textit{odgi draw} renders a 2D image of the graph (\S\ref{sec:viz}).
\item~\textit{odgi extract} excerpts subsets of the graph based on path ranges (\S\ref{sec:supp_edit}).
\item~\textit{odgi explode} breaks the graph into connected components (\S\ref{sec:supp_edit}).
\item~\textit{odgi squeeze} unifies disjoint graphs (\S\ref{sec:supp_edit}).
\item~\textit{odgi chop} breaks long nodes into shorter ones (\S\ref{sec:supp_edit}).
\item~\textit{odgi unchop} combines unitig nodes (\S\ref{sec:supp_edit}).
\item~\textit{odgi break} removes cycles in the graph (\S\ref{sec:supp_edit}).
\item~\textit{odgi prune} removes complex regions (\S\ref{sec:supp_edit}).
\item~\textit{odgi groom} resolves spurious inverting links (\S\ref{sec:supp_edit}).
\item~\textit{odgi position} lifts coordinates between path and graph positions (\S\ref{sec:untangle}).%
%\item \textit{odgi server} provides a HTTP server to lift coordinates (\S\ref{sec:untangle}).
\item~\textit{odgi untangle} deconvolutes paths relative to a reference (\S\ref{sec:untangle}).
\item~\textit{odgi tips} finds path end points relative to a reference (\S\ref{sec:supp_navigation}).
\item~\textit{odgi sort} orders the graph nodes (\S\ref{sec:sort}).
\item~\textit{odgi layout} establishes a 2D layout (\S\ref{sec:sort}).
\item~\textit{odgi matrix} derives the pangenome matrix (\S\ref{sec:supp_metrics}).
\item~\textit{odgi paths} lists and extracts paths in FASTA (\S\ref{sec:supp_metrics}).
\item~\textit{odgi flatten} converts the graph to FASTA and BED (\S\ref{sec:supp_metrics}).
\item~\textit{odgi stats} provides numerical properties of the graph (\S\ref{sec:metrics}).
\item~\textit{odgi bin} generates a summarized view of the graph (\S\ref{sec:supp_metrics}).
\item~\textit{odgi depth} describes node depth over graph and path positions (\S\ref{sec:metrics}).
\item~\textit{odgi degree} describes node degree over graph and path positions (\S\ref{sec:metrics}).
\end{itemize}

@@

\section{Supplementary}

@@

\subsection{Graph Editing}
\label{sec:supp_edit}
Subgraphs can be extracted by using the paths in the graph as coordinate systems to guide the process. For such operation, \textit{odgi extract} allows users to extract specific regions of the graph as defined by query criteria.
Regions of interest can be specified by graph nodes or path range(s), also in BED format. Furthermore, it is possible to indicate a list of paths to be preserved completely in the extracted graph.
We begin by collecting all graph nodes that fall within the ranges to extract (and the paths to preserve, if requested).
Users can specify the number of steps or nucleotides to expand the selection and include neighboring nodes.
Then, edges connecting all selected nodes are added in the subgraph under construction.
Finally, the portions of the paths (i.e., the subpaths) walking through the selected nodes are extracted and added to the new subgraph.
Subpaths are searched in parallel, created serially, and extended in parallel again thanks to the parallelism enabled by the ODGI data structure (see \S\ref{sec:build}), making \textit{odgi extract} a scalable solution to extract also complex subregions presenting nodes with very high node depth.

@@

\subsection{Graph Sorting}
\label{sec:supp_sorting}
Most subcommands in ODGI require and verify that the input graph’s node identifiers (IDs) are optimized, that is compacted from 1 to N where N is the number of nodes in the graph. 
If this assumption is violated, \textit{odgi sort} provides functionality to optimize the graph. 
This means that the first node identifier (ID) starts at 1 and the last node ID is the number of nodes. 
All sorting operations update the graph in place with an efficient ID rewriting algorithm.
The graph is then updated in place. 
First, the node identifiers are normalized (from 1 to number of nodes) including the adjustment of the edges. 
Second, path information, including both path metadata that points into the start and end steps of the path, plus each step of every path, is updated, too. 
We point out that changing the node order does not change our coordinate systems based on paths. 
These will now refer to a new node ordering.

@@

\subsection{Graph Metrics}
\label{sec:supp_metrics}
Pangenome graph topology can be derived by applying \textit{odgi matrix}, obtaining information on graph nodes connections in textual sparse matrix format.
To investigate on the genomic sequences encoded in the graph, \textit{odgi paths} allows users to calculate pairwise overlap statistics of groupings of paths and emit all path sequences in FASTA format, and it also allows the generation of a ``pangenome matrix'' that reports the copy number (presence/absence) of each path over each node.
\textit{odgi flatten} generates a linearization of the graph by emitting the pangenome sequence (the concatenation of all node sequences) in FASTA format, and the projection of all paths on the linearized sequence in BED format.
ODGI also offers more advanced tools for the interrogation of the graphs.
To study very large pangenomes, users can use \textit{odgi bin} to summarize the path information into bins of a specified size, generating a summarized view of gigabase scale graphs in TSV or JSON file formats.
This concept found its way into the pangenome graph ontology~\citep{Yokoyama2020}. %maybe we want to save this up for the discussion

>>>>

%\FIXME{Update the references to the sections.}
\begin{itemize}
\item~\textit{odgi build} constructs the ODGI data model from GFA file (\S\ref{sec:build}).
\item~\textit{odgi view} converts the ODGI data model into GFA file (\S\ref{sec:build}).
\item~\textit{odgi viz} provides a linear visualization of the graph (\S\ref{sec:viz}).
\item~\textit{odgi draw} renders a 2D image of the graph (\S\ref{sec:viz}).
\item~\textit{odgi extract} excerpts subsets of the graph based on path ranges (\S\ref{sec:supp_edit}).
\item~\textit{odgi explode} breaks the graph into connected components (\S\ref{sec:supp_edit}).
\item~\textit{odgi squeeze} unifies disjoint graphs (\S\ref{sec:supp_edit}).
\item~\textit{odgi chop} breaks long nodes into shorter ones (\S\ref{sec:supp_edit}).
\item~\textit{odgi unchop} combines unitig nodes (\S\ref{sec:supp_edit}).
\item~\textit{odgi break} removes cycles in the graph (\S\ref{sec:supp_edit}).
\item~\textit{odgi prune} removes complex regions (\S\ref{sec:supp_edit}).
\item~\textit{odgi groom} resolves spurious inverting links (\S\ref{sec:supp_edit}).
\item~\textit{odgi position} lifts coordinates between path and graph positions (\S\ref{sec:untangle}).%
%\item \textit{odgi server} provides a HTTP server to lift coordinates (\S\ref{sec:untangle}).
\item~\textit{odgi untangle} deconvolutes paths relative to a reference (\S\ref{sec:untangle}).
\item~\textit{odgi tips} finds path end points relative to a reference (\S\ref{sec:supp_navigation}).
\item~\textit{odgi sort} orders the graph nodes (\S\ref{sec:sort}).
\item~\textit{odgi layout} establishes a 2D layout (\S\ref{sec:sort}).
\item~\textit{odgi matrix} derives the pangenome matrix (\S\ref{sec:supp_metrics}).
\item~\textit{odgi paths} lists and extracts paths in FASTA (\S\ref{sec:supp_metrics}).
\item~\textit{odgi flatten} converts the graph to FASTA and BED (\S\ref{sec:supp_metrics}).
\REVIEWED{\item~\textit{odgi pav} generates a presence-abscence variation table (\S\ref{sec:supp_pav}).}
\item~\textit{odgi stats} provides numerical properties of the graph (\S\ref{sec:metrics}).
\item~\textit{odgi bin} generates a summarized view of the graph (\S\ref{sec:supp_metrics}).
\item~\textit{odgi depth} describes node depth over graph and path positions (\S\ref{sec:metrics}).
\item~\textit{odgi degree} describes node degree over graph and path positions (\S\ref{sec:metrics}).
\end{itemize}

@@

\section{Supplement}

@@

\subsection{Editing}
\label{sec:supp_edit}
Subgraphs can be extracted by using the paths in the graph as coordinate systems to guide the process. For such operation, \textit{odgi extract} allows users to extract specific regions of the graph as defined by query criteria.
Regions of interest can be specified by graph nodes or path range(s), also in BED format. Furthermore, it is possible to indicate a list of paths to be preserved completely in the extracted graph.
We begin by collecting all graph nodes that fall within the ranges to extract (and the paths to preserve, if requested).
Users can specify the number of steps or nucleotides to expand the selection and include neighboring nodes.
Then, edges connecting all selected nodes are added in the subgraph under construction.
Finally, the portions of the paths (i.e., the subpaths) walking through the selected nodes are extracted and added to the new subgraph.
Subpaths are searched in parallel, created serially, and extended in parallel again thanks to the parallelism enabled by the ODGI data structure (see \S\ref{sec:build}), making \textit{odgi extract} a scalable solution to extract also complex subregions presenting nodes with very high node depth.

@@

\subsection{Sorting and node identifier compaction}
\label{sec:supp_sorting}
Most subcommands in ODGI require and verify that the input graph’s node identifiers (IDs) are optimized, that is compacted from 1 to N where N is the number of nodes in the graph. 
If this assumption is violated, \textit{odgi sort} provides functionality to optimize the graph. 
This means that the first node identifier (ID) starts at 1 and the last node ID is the number of nodes. 
All sorting operations update the graph in place with an efficient ID rewriting algorithm.
The graph is then updated in place. 
First, the node identifiers are normalized (from 1 to number of nodes) including the adjustment of the edges. 
Second, path information, including both path metadata that points into the start and end steps of the path, plus each step of every path, is updated, too. 
We point out that changing the node order does not change our coordinate systems based on paths. 
These will now refer to a new node ordering.

@@

\subsection{Linear projections}
\label{sec:supp_metrics}
Pangenome graph topology can be derived by applying \textit{odgi matrix}, obtaining information on graph nodes connections in textual sparse matrix format.
To investigate on the genomic sequences encoded in the graph, \textit{odgi paths} allows users to calculate pairwise overlap statistics of groupings of paths and emit all path sequences in FASTA format, and it also allows the generation of a ``pangenome matrix'' that reports the copy number (presence/absence) of each path over each node.

In standard practice, pangenome analysis examines presence-absence variations (PAVs), which correspond to loci that are present in some samples but not others.
\textit{odgi pav} uses a set of genomic intervals (in BED expressed in the coordinate space of paths in the graph) to demarcate PAVs.
It then describes the coverage of all paths relative to these PAVs, yielding matrix or tabular representation.
The precise determination of PAVs based on graph topology remains an open problem, but practically any method capable of generating a BED file can be used here.
This lets us define PAVs using repeat or gene annotations of the genomes represented as paths in the graph.
A simple technique is to take the output of \textit{odgi flatten}, which generates a linearization of the graph by emitting the pangenome sequence (the concatenation of all node sequences) in FASTA format, and the projection of all paths on the linearized sequence in BED format relative to the graph's paths.

ODGI also supports an experimental binned representation of graphs designed to support the study and visualization of pangenomes at different scales of resolution.
\textit{odgi bin} summarizes graph path information into bins of a specified size, generating a summarized view of gigabase scale graphs in TSV or JSON file formats.
We have further supported this binning approach in pangenome graph ontology model~\citep{Yokoyama2020}. %maybe we want to save this up for the discussion
